Avoiding the Global Sort: \break A Faster Contour Tree Algorithm: Erratum

### 2021-09-23 erratum/updated/erratum.tex
updated erratum

\author{
  Benjamin Raichel\thanks{%
      Department of Computer Science; %
      University of Texas at Dallas; %
      Richardson, TX, 75080, USA; %
      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}}.
}
  \and
  C. Seshadhri\thanks{%
        Department of Computer Science; %
      University of California, Santa Cruz; %
      Santa Cruz, CA, 95064, USA; %
      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
  }
}

\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
% \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
}
\date{}

@@

\begin{abstract}
We revisit the classical problem of computing the \emph{contour tree} of a scalar field $f:\MM \to \RR$, where $\MM$ is a triangulations of a ball in $\RR^d$. The contour tree is a fundamental topological structure that tracks the evolution of level sets of $f$ and has numerous applications in data analysis and visualization.

All existing algorithms begin with a global sort of at least all critical values of $f$, which can require (roughly) $\Omega(n\log n)$ time, where $n$ is the number of vertices of the mesh. Existing lower bounds show that there are pathological instances where this sort is required. We present the first algorithm whose time complexity depends on the contour tree structure, and avoids the global sort for non-pathological inputs. (We assume that the Morse degree of the function $f$ at any point is at most $3$.) If $C$ denotes the set of critical points in $\MM$, the running time is roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$. This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat. Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ or in the same descending path in the contour tree, allowing us to argue strong optimality properties of our algorithm.

Our algorithm requires several novel ideas: partitioning $\MM$ in well-behaved portions, a local growing procedure to iteratively build contour trees, and the use of heavy path decompositions for the time complexity analysis. 
\end{abstract}

\newpage
\pagenumbering{arabic}

\section{Introduction}

Geometric data is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a topological structure used to represent and visualize
the function $f$. 
It is convenient to think of $f$ as a simplicial complex sitting in $\RR^{d+1}$,
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components. 
The contour tree is a concise representation of these events \cite{kobps-ctsssit-97,csa-cctad-03}.
%Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
%For $d>2$, some subsequent works, such as \cite{csa-cctad-03}, only include changes in the number of components.

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes. The contour tree tracks a specific
subset of events, called ``joins" and ``splits".
We provide formal definitions later.
An edge of the contour tree connects two such critical points if one event immediately ``follows"
the other as the sweep plane makes its pass.  
\Fig{relative} and \Fig{sorting} 
show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards. Leaves are either maxima or minima,
and internal nodes are either ``joins" or ``splits".

% so we can talk of the ``up-degree" and ``down-degree" of vertices.
% Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
% $1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
% merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
% In dimensions $d \geq 3$, we may have internal vertices of both up and down-degree $1$. 
% For example, these could be events where contours gain or lose handles.

Consider $f: \MM \to \RR$, where $\MM$ is the triangulation of a $d$-dimensional ball with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$, see \Def{deg}. 
The degree assumption can be achieved via vertex unfolding, as discussed later,
and is standard in this literature~\cite{kobps-ctsssit-97}.)
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

All previous algorithms begin by sorting (at least) the critical points. Can we beat this sorting bound for certain instances,
and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
do not need to be compared. Look at \Fig{relative} to see such an example. All previous algorithms waste time
sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
trees joined at their roots, and we do not need the entire sorted order of critical points
to construct the contour tree.

\begin{figure}[h]\centering
    \includegraphics[width=.23\linewidth]{figs/spikey1}%
    \hspace{.37in}
    \includegraphics[width=.23\linewidth]{figs/spikey2}%
    \hspace{.41in}
    \includegraphics[width=.21\linewidth]{figs/spikeyTree}
    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
    \label{fig:relative}
\end{figure}

\begin{figure}[h]\centering
    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced4}
    \hspace{.3in}
    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
    \caption{On left, a surface with a balanced contour tree, but whose join and split trees have long tails.  
    On right (from left to right), the contour, join and split trees.}
    \label{fig:sorting}
\end{figure}

Our main result gives an affirmative answer. Remember that we can consider
the contour tree as directed from top to bottom. For any node $v$ in the tree, let $\ell_v$ denote the length of the 
longest directed path passing through $v$.

\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, where $\MM$
is a triangulation of a $d$-dimensional ball.
Let $N$ be the number of facets in $\MM$.
Let the Morse degree of $f$ at any point be at most $3$.
Denote
the contour tree by $T$ with vertex set (the critical points) $C(T)$. 
There exists an algorithm to compute $T$
in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
compares function values at pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

Essentially, the ``run time per critical point" is determined by the height/depth of the point in the contour tree.
This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
where $D$ is the directed diameter of the contour tree. 
This is a significant asymptotic improvement for short and fat contour trees. For example, if the tree is balanced,
then we get a bound of $O(t\log\log t)$.
Even if $T$ contains a long path of length $O(t/\log t)$, but is otherwise short, we get the improved bound of $O(t\log\log t)$.

\subsection{A refined bound with optimality properties}\label{sec:more-refined}

\Thm{main-corr} is a direct corollary of a stronger but more cumbersome theorem.

\begin{definition}
\label{def:path} 
For a contour tree $T$, a \emph{leaf path} is any path in $T$ containing a leaf, 
which is also monotonic in the height values of its vertices. 
Then a \emph{path decomposition}, $P(T)$, is a partition of the vertices of $T$ into a set of vertex disjoint leaf paths.  
\end{definition}

\begin{theorem} 
\label{thm:main-alg} 
(Consider the input setting of \Thm{main-corr}.)
There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
In particular, comparisons are only made between pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

Note that \Thm{main-corr} is a direct corollary of this statement. For any $v$, $\ell_v$ is at least
the length of the path in $P(T)$ that contains $v$.
This bound is strictly stronger,
since for any balanced contour tree, the run time bound of \Thm{main-alg} is $O(t\alpha(t) + N)$, and $O(N)$
comparisons are made. (Since for a balanced tree, one can show $\sum_{p \in P(T)} |p|\log |p|= O(t)$.)

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value $t=\Omega(N)$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $c t\log t$, for some constant $c$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $\alpha \thmConst$ comparisons, for any constant $\alpha$. Then \emph{any algorithm} must make roughly $\thmConst$
comparisons in the worst-case over this set. (Further details are in \Sec{lb}.)

\begin{theorem} 
\label{thm:main-lb}
There exists some absolute constant $\alpha$ such that the following holds.
For sufficiently large $t$ and any $\thmConst\in [t, t\log t]$, consider the set $\bF_\thmConst$ of terrains with $t$ critical points such that
the number of comparisons made by the algorithm of \Thm{main-alg} on these terrains is in $[\thmConst,\alpha \thmConst]$.
Any algebraic decision tree that correctly computes
the contour tree on all of $\bF_\thmConst$ has a worst case running time of $\Omega(\thmConst)$.
\end{theorem}
% 
% \myparagraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
% of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
% characterize hard instances.
% % explain \emph{when}
% % computing contour trees is expensive, and provide insight into the hardness of particular instances. 
% %It is intriguing
% %that the cost (as defined earlier) roughly captures the hardness of an instance.
% Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
% This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
% Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
% of easy to hard instances. In general, our analysis forces one to look closely
% at the relationship between sorting and the contour tree problem. 
%

\subsection{Previous work}

Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
Contour trees have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour tree definitions and algorithms.

The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalyi \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-03} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.

Contour trees are a special case of Reeb graphs, a general topological representation for real-valued functions
on any manifold. Algorithms for computing Reeb graphs
is an active topic of research~\cite{sk-crgacs-91,cehnp-lrbm-03,PaScBr07,DoNa09,HaWaWe10,Pa12}, where
two results explicitly reduce to computing contour trees~\cite{TiGySi09,DoNa13}.

% \paragraph{Organization.}  
% \Sec{basics} contains basic definition for contour trees.  In \Sec{approach} we give a high level overview and intuition for our approach 
% for computing contour trees.  \Sec{surgery} and \Sec{rain} describe how we break the intial input into nicely behaved (extemum dominant) pieces. 
% \Sec{extreme} shows that computing the contour tree for these nicely behaved pieces is roughly equivalent to computing the join tree. 
% \Sec{paint} contains our algorithm for computing join trees, which along with previous sections gives our algorithm for contour trees.  
% Finally, in \Sec{pathDecomp} and \Sec{lb} we relate the running time of our algorithm (providing upper and lower bounds) to a particular decomposition of the contour tree.

\section{Contour tree basics} \label{sec:basics}

We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a fully triangulated simplicial complex of a topological ball in $\RR^d$,
except for specially designated \emph{boundary facets}. We assume that all pairs of vertices from $\MM$ have distinct function values, except for pairs from the same boundary facet,
and that $f$ is only explicitly defined on the vertices, and all other values are obtained by linear interpolation.

We assume that the boundary values satisfy a special property. This is mainly for convenience
in presentation.

\begin{definition} \label{def:bound} The function $f$ is \emph{boundary critical} if the following holds.
Consider a boundary facet $F$. All vertices of $F$ have the same function value. Furthermore, all
neighbors of vertices in $F$, which are not also in $F$ itself, 
either have all function values strictly greater than or all function values strictly less than the function value at $F$.
\end{definition}

Boundary facets allow us to capture the resulting surface pieces after our algorithm makes a horizontal cut. 
The above definition is convenient, as any point within a given facet has a well-defined height, including the boundary facets.

We think of the dimension $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time access to neighboring simplices
in $\MM$ (e.g. incidence graphs~\cite{Ed87}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$, 
where $f(x)$ is the ``height" of a point $x \in \MM$, which is encoded in the representation of $\MM$. 
Specifically, rather than writing our input as $(\MM,f)$, we abuse notation and typically just write $\MM$ to denote the lifted complex.

\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x\in \MM| f(x) = h\}$.
A \emph{contour} is a connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
\end{definition}

Note that a contour that does not contain a boundary is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
We let $\delta$ and $\eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
$f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.
The following definition of a critical point is not standard,
but is convenient for our presentation.

\begin{definition} \label{def:deg} The \emph{Morse up-degree} of $x$ is the number of $(f(x) + \delta)$-contours of $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$. The \emph{Morse down-degree} is the number of $(f(x) - \delta)$-contours of $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$.

A \emph{regular} point has both Morse up-degree and down-degree $1$. A \emph{maximum} has Morse up-degree $0$, while
a \emph{minimum} has Morse down-degree $0$. A \emph{Morse Join} has Morse up-degree strictly greater than $1$,
while a \emph{Morse Split} has Morse down-degree strictly greater than $1$. Non-regular points are called \emph{critical}.
\end{definition}

The set of critical points is denoted by $\cV(f)$.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called \emph{critical}, if $f(v) = h$, for some $v \in \cV(f)$. 
A contour is called \emph{critical}, if it contains a critical point, and it is called \emph{regular} otherwise.

The critical points are where certain topological properties
 of level sets change.
By assuming that our simplicial complex is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maximum or minimum.

\begin{definition} \label{def:equiv} Two regular contours $\psi$ and $\psi'$ are \emph{equivalent} 
if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
such that no $x \in p$ belongs to a critical contour.
\end{definition}

This equivalence relation gives a set of \emph{contour classes}. 
Every such class maps to intervals
of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
to be created at $x_i$ and destroyed at $x_j$.

\begin{definition} \label{def:tree} The \emph{contour tree} is the graph on vertex set $\cV= \cV(f)$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher to lower function value.)
\end{definition}

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternative definitions of the contour tree
in~\cite{csa-cctad-03} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.

We assume that the Morse up (or down) degree of any point
in $\MM$ is at most $2$, and that the total Morse degree
is at most $3$. This immediately implies analogous bounds
on the up and down-degrees in $\reeb(\MM)$. This
is standard in the literature
among others, though it can be challenging to enforce in 
practice. Theoretically, a manifold with points of higher Morse degree
can be converted to one with the degree condition through
a process of \emph{vertex unfolding} (refer to
Section 6.3 in~\cite{HaEd10}). This effectively converts
a multi-saddle to a set of ordinary saddles.
%A detailed discussion is provided in the end
%of Section 2 and beginning of Section 3 in~\cite{BaKr+98}.
For 2D manifolds,
this conversion does not increase the surface complexity.
In higher dimensions, the unfolding can increase complexity if
$\MM$ has vertices of non-constant degree.

% Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
% via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
% Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
%
%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Some remarks}
{\bf 1)}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.\\

\noindent{\bf 2)}
As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every regular vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees.
\Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{A tour of the new contour tree algorithm} \label{sec:approach}
We provide a high-level description of the main ideas of subsequent sections.

\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
with exactly the same contour tree, but different orderings of (heights of) the critical points.
Turning it around, we cannot deduce the full height ordering of critical points from the contour tree, and so 
sorting all critical points is computationally unnecessary.
In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
another of the splits. Again, it is not necessary to know the relative ordering between the mounds
on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
information is necessary: on the left, the little valleys are higher than the big central valley,
and this is reflected in the contour tree. Leaf paths in the contour tree have points
in sorted order, but incomparable points in the tree are unconstrained.
How do we sort exactly what is required, without knowing the contour tree in advance?

\myparagraph{Join and split trees:}
The standard contour tree algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-03}, builds
two different trees, called the join and split trees, and then merges them together to get the contour tree.
Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree, $\cJ(\MM)$, tracks exactly these events. 
Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
vertices higher than $v$. 
Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
are still put as vertices in the join tree. The split tree
is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.

A major insight of~\cite{csa-cctad-03} is an ingeniously simple linear time procedure to construct
the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-03}
to get the contour tree requires having the split vertices in the join tree. Without
this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.

Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
of the pieces is linear (see \Sec{rain} for details).

We now have carved $\MM$ up into a collection of extremum dominant pieces. 
In \Sec{surgery} we prove a simple contour surgery theorem allowing us to build the contour tree of $\MM$
from the contour trees of these various pieces.  Specifically, we argue our cutting in $\MM$ was done in such a way that 
each cut corresponds to cutting only a single edge in the contour tree.  Informally this means gluing $\MM$ back together along these 
cuts corresponds to gluing the contour tree together.

Ultimately, the reason for cutting $\MM$ into extremum dominant pieces is that such pieces have simpler contour tree structure. 
Specifically, using ideas from~\cite{csa-cctad-03}, in \Sec{extreme} we prove that the contour tree of an extremum dominant complex is (basically) just the join tree. 
Thus there is no longer a disconnect between the amount of sorting required for the contour tree versus the join tree, 
and so the problem of computing contour trees efficiently reduces to that for join trees.

\myparagraph{Join trees from painted mountaintops:} Our main result is a faster algorithm for join trees.
The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
colors for each maximum. In arbitrary order, spill paint from each maximum, wait till it flows down,
then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
paint the interior of faces with dimension strictly larger than 1.} 
Furthermore, paints do not mix, so each edge receives a unique color, decided by the first paint to reach it.  
%
Note that the raining procedure, used above to carve the input into extremum dominant pieces, 
cannot be used here as it wets interiors of higher dimensional faces and so raining from each maximum may significantly increase the input complexity (see \Fig{order}).
On the other hand, the interface between two different colors is \emph{not} a contour, and so the join trees of each color class cannot just be simply glued together.

% In the following, $[n]$ denotes $\{1, \ldots, n\}$, for any natural number $n$.
% \begin{definition} \label{def:paint1} \InSoCGVer{(Restatement of \Def{paint2}.)} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
% A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
%  Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
% 	consisting of edges in $E$, such that all edges along this path have the same color as $x$. 
% 
% An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
% \end{definition}

% Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
% of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
% colors are distinct. A few comments on paint vs water. The interface between two regions of different color
% is \emph{not} a contour, and so we cannot apply the divide-and-conquer
% approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
% Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
% Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
% dominant input, using water to divide the input $\NN$ raises the complexity too much.

Our algorithm incrementally builds $\cJ(\MM)$ from the leaves (maxima) to the root (dominant minimum), by repeatedly merging colors and components together.  
We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
Let us focus on the initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
Both mounds are colored completely blue or red, while $v$ is touched by both colors.
So this indicates that $v$ joins the blue maximum and red maximum in $\cJ(\MM)$.

This is precisely how we hope to exploit the information in the painting. We prove later that
when some join $v$ has exactly two colors among all incident edges with $v$ as the lower endpoint, the corresponding maxima (of those colors) are exactly
the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
is now a new maximum, from which purple paint is poured. In terms of $\cJ(\MM)$, this
is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
Alternatively, $\cJ(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
data structure for that. So the red and blue mounds are not actually recolored purple, 
and instead say red now points to blue in the union find data structure.

>>>>

\author{
  Benjamin Raichel\thanks{%
      Department of Computer Science; %
      University of Texas at Dallas; %
      Richardson, TX, 75080, USA; %
      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}}.
}
  \and
  C. Seshadhri\thanks{%
        Department of Computer Science; %
      University of California, Santa Cruz; %
      Santa Cruz, CA, 95064, USA; %
      \texttt{\si{sesh}@\si{ucsc}.\si{edu}} %
  }
}

\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm: Erratum}
% \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%

\date{}

@@

\begin{abstract}
This erratum points out a serious bug in our paper, ``Avoiding the Global Sort: A Faster Contour Tree Algorithm"~\cite{rs-17},
as pointed out by Rote~\cite{rote}.

\end{abstract}

\pagenumbering{arabic}

The paper~\cite{rs-17} presents a new algorithm to compute contour trees, whose running time depends on the shape of the tree, and in particular avoids the initial global sort of all the critical points.  The algorithm has two parts. The first part 
is a new algorithm to compute join trees, again whose running time depends on the shape of the tree. The second part reduces computing the contour tree to join tree computations, by cutting the input simplicial complex into disjoint pieces. For each of these pieces, the contour tree and join tree are the same. The paper claims this decomposition can be performed in linear time. 
As pointed out by Rote, the proof of this claim is incorrect~\cite{rote}. As a result, the main theorem (Theorem 1.1) does not hold.

The problems appear in Section 6 of~\cite{rs-17}. 
For a maximum $x\in \MM$, as defined in Section 6, $\wet(x,\MM)$ is the set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$. A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$ has non-trivial intersection with $\wet(x,\MM)$ (i.e.\ the intersection is neither empty nor the entire neighborhood). 
The bug is that the paper does not describe precisely how to compute $\wet(x,\MM)$. (This appears in the first step of the procedure {\tt rain}$(x,\NN)$.) The natural approach is to perform a descending BFS or DFS from $x$. Suppose we reach a critical vertex $v$ which is a join. We wet an edge in one of its up-stars when we reached it, and its other up-star is entirely dry since it is a join. As our procedure wets the interior of faces, at this point we naturally wish to walk around the interface contour adjacent to this join, and then rain down (i.e. continue the DFS) from the contour. 
%
For two distinct join vertices, with a wet edge in one up-star, the edge sets crossed by their corresponding dry contours will be disjoint.  
%
Thus the cost of walking along these contours for all joins can be charged to the size of the interface which is linear in the size of $\wet(x,\MM)$. 
%
The issue, however, is that while we know locally whether a given vertex $v$ is a critical vertex, we do not know whether it is a join.
%
So suppose $v$ is critical with one dry up-star, but that it is not a join. Then one can still walk around the contour from the dry up-star and rain down. However, if we do this for all such $v$ it may no longer be true that the edge sets crossed by the contours are disjoint. Thus depending on the order in which vertices are handled, this could lead to edges being traversed a super-constant number of times.

In summary, while the actual cutting can be performed in linear time, it is not clear how long it takes to determine where to make the cuts. Specifically, it is not clear how long it takes to determine the interface of the wet complex, namely the first step in the {\tt rain} procedure of section Section 6, and thus it is unclear whether Theorem 6.7 holds.
At the most basic level, the issue boils down to efficiently determining which critical points are joins. There does not exist an efficient procedure
for finding all joins.

We note that that the algorithm and analysis of computing the join tree still hold (Sections 4 and 8 of~\cite{rs-17}) .
